---
title: "The Role of Personal Entry Characteristics in Dropout from Teacher Education"
subtitle: "Doctoral Colloquium, 21 February 2024" 
title-block-banner: true
author: David Simon [david.simon@dipf.de]
date: today
date-format: "DD MMMM YYYY"
editor: source
execute:
  echo: true
  warning: false
  message: false
  cache: false
format: 
  html:
    theme: flatly
    fontsize: 1.0em
    toc: true
    toc-location: left
    toc-depth: 4
    embed-resources: true
    code-fold: show
    code-tools: true
    code-link: true
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl
bibliography: references.bib
---

## Introduction

Extended summary (parts) proposal SIG 11 & 13 Conference 2024, tba.
Example reference [@heublein2014].
The preregistrationis avaliable under https://doi.org/10.17605/OSF.IO/D2K7Y

## Analysis

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this [@heublein2014]:

### Step 1

```{r}
#| echo: false
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).

## 50 Topics to Spark Good Conversation
- Question 1
